---
title: "R Notebook"
output: html_notebook
---

Load the packages
```{r}
library(tidyverse)#wrangle data
library(dplyr) #cleandata
library(lubridate) #wrangle data attributes
library(ggplot2) #visualize data
install.packages("readr")
library(readr)
install.packages("skimr")
library(skimr)
```

STEP 1 : prepare the data and comn=bine them in one data frame
```{r}
getwd()
setwd("/Users/macbookpro/Downloads/CYCLISTIC")

```

STEP 2 : combine the data from september 2020 to August 2021 into one data frame
```{r}
bike_data <- rbind(
  read_csv("202009-divvy-tripdata.csv"),
  read_csv("202010-divvy-tripdata.csv"),
  read_csv("202011-divvy-tripdata.csv"),
  read_csv("202012-divvy-tripdata.csv"),
  read_csv("202101-divvy-tripdata.csv"),
  read_csv("202102-divvy-tripdata.csv"),
  read_csv("202103-divvy-tripdata.csv"),
  read_csv("202104-divvy-tripdata.csv"),
  read_csv("202105-divvy-tripdata.csv"),
  read_csv("202106-divvy-tripdata.csv"),
  read_csv("202107-divvy-tripdata.csv"),
  read_csv("202108-divvy-tripdata.csv")
)
```

STEP 3 : Examine the data frame
```{r}
head(bike_data)  #to get the first few records from all the columns displayed
dim(bike_data) #to show the number of rows and columns the data frame has
colnames(bike_data) #to identify all the colums in the bike_data 
summary(bike_data) #to generate descriptive statistics of a data object e.g mean, median
```

STEP 4 :check for the null values(NA)
```{r}
sum(is.na(bike_data)) #choosing not to remove NA because most NA are related to the start and end station names
```

STEP 5:drop the columns we don't need : start_lat, start_lng, end_lat, end_lng
```{r}
bike_data <- bike_data %>%
  select(-c(start_lat,start_lng,end_lat,end_lng))
colnames(bike_data)
```

STEP 6:Add column ride_length which is the lenght of each ride from ended_at minus started_at and format HH:MM:SS,Add also column day_of_the_week and calculate the day of the week that each ride started.
```{r}
bike_data <- bike_data %>%
  mutate(ride_length = ended_at - started_at) %>%
  mutate(day_of_the_week = weekdays(as.Date(bike_data$started_at)))

```
```{r}
#convert ride_length from seconds into minutes
bike_data$ride_length <- as.numeric(bike_data$ride_length)
bike_data$ride_length <- as.numeric(bike_data$ride_length/60)
head(bike_data)

```

STEP 7:removing the bad data and doing analysis on the ride_length column
check for the data with negative ride_length
```{r}
bike_data <- bike_data[bike_data$ride_length>0,]
```

```{r}
#check for data with ride_length more than 1 day (1440minutes).3484 rides went down for more than one day
sum(bike_data$ride_length > 1440)

```

```{r}
#check for extreme ontliner:55944.15 minutes or 38.83 daysfor the max ride and less than 1sec for the least ride 
min(bike_data$ride_length)
max(bike_data$ride_length)
```

```{r}
#check the mean and median
mean(bike_data$ride_length)
median(bike_data$ride_length)
```

STEP 8:Aggregate to analyze the data based on user type: member vs casual
```{r}
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = mean)
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = median)

```

```{r}
#Average ride time by each day for memebers vs casuals
bike_data$day_of_the_week <- ordered(bike_data$day_of_the_week, levels=c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday"))
aggregate(bike_data$ride_length ~ bike_data$member_casual + bike_data$day_of_the_week, FUN = mean)
```

STEP 9:Further Analysis into the stations, which shows where NA belongs in step 5
```{r}
head(count(bike_data, start_station_name, member_casual, rideable_type, sort = TRUE))
head(count(bike_data, end_station_name, member_casual, rideable_type, sort = TRUE))
```

STEP 10:analyze ridership by type and weekday
```{r}
bike_data %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>% #creates weekday field using wday()
  group_by(member_casual, weekday) %>% #groups by usertype and weekday
  summarise(number_of_rides = n(),          #calculates the number of rides
            average_duration = mean(ride_length)) %>% #calculate the average duration
  arrange(member_casual, weekday)     #sorts
```


STEP 11:visualize the number of rides by rider types and average duration
```{r}
par(mfrow=c(2,2)) ##create a grid of plots with 2 rows and 2 columns.

bike_data %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual,weekday)%>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill= member_casual))+
  geom_col(position = "dodge")
```

STEP 12:visualize memeber  vs casual on short ride (less and one hour)
```{r}
one_hour_data <- bike_data %>% filter(ride_length < 60)
one_hour_data$ride_length_by_mins <- cut(one_hour_data$ride_length, breaks = 20)

ggplot(data = one_hour_data)+
  geom_bar(mapping = aes(x = ride_length_by_mins, fill= member_casual)) +
  labs(title = "one hour ride length") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1))+
  scale_x_discrete(labels=c("3", "6","9","12", "15","18","21","24","27","30","33","39","42","45","48","51","54","57","60"))

```

STEP 13:visualize day of the week riding choices between member vs casual
```{r}
ggplot(data = bike_data) + 
  geom_bar(mapping = aes(x = factor(day_of_the_week),fill= rideable_type))+
  facet_wrap(~member_casual)+
  labs(title = 'riding choice during the day of the week', x= 'day of the week')+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

STEP 14:check for peak time for bike usage between member vs casual
```{r}
hour_data <- bike_data
hour_data$start_hour <- as.numeric(format(strptime(bike_data$started_at,"%Y-%m-%d %H:%M:%OS"),'%H'))

ggplot(data = hour_data) +
  geom_bar(mapping = aes(x=start_hour, fill = member_casual), stat = 'count')+
  facet_wrap(~factor(day_of_the_week))+
  labs(title = "bike usage by starting hour", x="starting hour")+
  theme(axis.text.x = element_text(angle= 90, vjust = 0.5, hjust = 1))

```

STEP 15:save as csv for further analysis and visualization in tableau,the data frame
```{r}
write_csv(bike_data,"bikedata.csv")

```

```{r}
#total and average weekly rides by rider type 
summary_ride_weekly <- bike_data %>%
  mutate(weekday = wday(started_at, label = TRUE )) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length))%>%
  arrange(member_casual, weekday)

write_csv(summary_ride_weekly,"summary_ride_weekly.csv")
```

```{r}
#total and average weekly rides by rider type
summary_ride_weekly <- bike_data %>%
  mutate(weekday = wday(started_at, label = TRUE )) %>%
  group_by(member_casual, weekday, rideable_type) %>%
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length))%>%
  arrange(member_casual, weekday)

write_csv(summary_ride_weekly,"summary_ride_weekly_type.csv")
```

```{r}
#total and average monthly rides by rider type
summary_month <- bike_data %>%
  mutate(month = month(started_at, label = TRUE )) %>%
  group_by(month,member_casual) %>%
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length))%>%
  arrange(month,member_casual)

write_csv(summary_ride_weekly,"summary_ride_monthly.csv")
```

```{r}
#most popular stations
popular_stations <- bike_data %>%
  mutate(station = start_station_name) %>%
  drop_na(start_station_name) %>%
  group_by(start_station_name, member_casual) %>%
  summarise(number_of_rides=n())

write.csv(popular_stations, "popular_stations.csv")
```

```{r}
#total memebership types and rideable types 
total_riders <- data.frame(table(bike_data$member_casual))
total_types <- data.frame(table(bike_data$rideable_type))

write_csv(total_riders, "total_riders.csv")
write_csv(total_types, "total_types.csv")

```



